---
title: Analysing Critic Scores as a Predictor of Video Game Sales
author: 
- Jaan Tollander de Balsch
- Samuel Piirainen
data: ""
output: 
  pdf_document: 
    latex_engine: xelatex
    toc: yes
    toc_depth: 1
bibliography: bibliography.bib
urlcolor: blue
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/harvard-anglia-ruskin-university.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import the required libraries and set Stan options.
```{r, message = FALSE}
library(ggplot2)
library(analogue)
library(tidyr)
library(plyr)
library(dplyr)
library(rstan)
library(loo)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


# The dataset
The dataset consists of *video games sales data* as of December 2016 scraped from Metacritic by @video_games_sales. In the dataset, each row represents data for a single game. The relevant fields for each row in the dataset are

* `Platform`: The abbreviation of the platform for playing the game.
* `Year_of_Release`: The year that the game was released. It contains games released between the years 1985 and 2016.
* `Global_Sales`: The number of game sales in millions.
* `Critic_Score`: A score given by critics. A number between 0 and 100.

In this analysis, we are interested in games released between the years 2000 - 2014. Also, we focused on the most popular platform in our analysis and filtered the less popular ones. 

We filtered out games from the years 2015 and 2016 because at the time when the data was gathered, these games were less than two years old, therefore they might sell more copies and the sales numbers would not represent the actual number of sales.

```{r}
table <- read.csv(
  file="Video_Games_Sales_as_at_22_Dec_2016.csv", 
  header=TRUE, sep=",", fill=TRUE, stringsAsFactors=FALSE)
```

```{r}
data <- filter(table, Year_of_Release >= 2009, Platform != "GBA", Platform != "GC", Platform != "PS")
data_scored = drop_na(
  data, Year_of_Release, Platform, Global_Sales, Critic_Score)
```

This plot visualizes the total global sales per year.
```{r}
d1 = group_by(data, Year_of_Release) %>%
  summarize(sales = sum(Global_Sales))
s1 = group_by(data_scored, Year_of_Release) %>%
  summarize(sales = sum(Global_Sales))
ggplot() + 
  geom_col(aes(x = Year_of_Release, y = sales), d1) +
  geom_col(aes(x = Year_of_Release, y = sales), s1, fill = "red") +
  labs(x = "Year of release", y = "Total global sales")
```

This plot visualizes the total global sales per platform.
```{r}
d2 = group_by(data, Platform) %>%
  summarize(sales = sum(Global_Sales))
s2 = group_by(data_scored, Platform) %>%
  summarize(sales = sum(Global_Sales))
ggplot() +
  geom_col(aes(x = reorder(Platform, -sales), y = sales), d2) +
  geom_col(aes(x = reorder(Platform, -sales), y = sales), s2, fill = "red") +
  labs(x = "Platform", y = "Total global sales")
```

This plot visualizes the total number of games per platform.
```{r}
d3 = count(data, Platform)
s3 = count(data_scored, Platform)
ggplot() + 
  geom_col(aes(x = reorder(Platform, -n), y = n), d3) +
  geom_col(aes(x = reorder(Platform, -n), y = n), s3, fill = "red") +
  labs(x = "Platform", y = "Total number of games") +
  theme(legend.position = "right")
```

The data indicates that the critics have rated a large percentage of games by total sales, but a smaller percentage by the total number of games. Therefore, critics have not rated many unpopular games.

This histogram visualizes counts of the critic scores.
```{r}
df <- data.frame(critic_score=data_scored$Critic_Score)
ggplot(df, aes(x=critic_score)) + geom_histogram(binwidth=2) + xlab("critic scores")
```

As can be seen from the critic scores, there are far less scores below $50$ than above $50$.


# The analysis problem
Our analysis problem is the question of whether critic scores predict global sales of video games. The critic scores and the sales numbers are assumed to be independent, which means that we assume that the critics have scored the games without the knowledge of their sales. In reality, this assumption might not always be valid. Even if the critics score the game before the publisher has sold any copies, the knowledge about the game's predecessor or the publisher could bias the critic's score.

# Description of the models
We use three models in this report; a separate, pooled and a hierarchical model. In the separate model each platform is represented by a separate model with its own standard deviation $\sigma_j$. In the pooled model we treat all platforms as equal. The whole datasets shares a single mean $\mu$ and standard deviation $\sigma$. In the hierarchical model we formulate a model where each group represents a single video game platform. It uses a shared standard deviation $\sigma$ between all the platforms.

Each model finds fits for both video game sales and critic scores. We removed unpopular platforms, `N64`, `DC`, `GB`, `WS`, from this analysis.

```{r}
data_models <- summarise(drop_na(group_by(filter(table, Year_of_Release >= 2009, Platform != "GBA", Platform != "GC", Platform != "PS"), Platform, Name, Year_of_Release, Global_Sales, Critic_Score, Rating)))
```

## Stan model prior choices

We give the scores an adjustable prior, because it has a scale that is common to each platform (Between 0 and 100).

We choose an informative prior mean of 70 for the scores, because this is reasonably close to the sample mean for Critic_Score of `r mean(data_models$Critic_Score)` (also see histogram above). We set the prior standard deviation to 0.2 to represent a slight amount of ignorance. We are saying that our prior guess for the critic score mean could be off by a factor of $\exp(0.2)=1.2$. (source: http://www.stat.columbia.edu/~gelman/research/unpublished/objectivityr3.pdf slide 13)

```{stan output.var="sales_separate"}
data {
  int<lower=0> N; // number of data points
  int<lower=0> K; // number of groups
  int<lower=1,upper=K> x[N]; // group indicator
  vector[N] sales; // sales
  vector[N] scores; // critic scores
  real pmu_scores; // prior mean
  real psigma_scores; // prior std
}
parameters {
  vector[K] mu_sales;
  vector[K] mu_scores;
  real<lower=0> sigma_sales[K];
  real<lower=0> sigma_scores[K];
}
model {
  mu_scores ~ normal(pmu_scores, psigma_scores);
  sales ~ normal(mu_sales[x], sigma_sales[x]);
  scores ~ normal(mu_scores[x], sigma_scores[x]);
}
generated quantities {
  vector[N] sales_log_lik;
  vector[N] scores_log_lik;
  for (i in 1:N) {
    sales_log_lik[i] = normal_lpdf(sales[i] | mu_sales[x[i]], sigma_sales[x[i]]);
  }
  for (i in 1:N) {
    scores_log_lik[i] = normal_lpdf(scores[i] | mu_scores[x[i]], sigma_scores[x[i]]);
  }
}
```

```{stan output.var="sales_pooled"}
data {
  int<lower=0> N; // number of data points
  vector[N] sales; // sales
  vector[N] scores; // critic scores
  real pmu_scores; // prior mean
  real psigma_scores; // prior std
}
parameters {
  real mu_sales;
  real mu_scores;
  real<lower=0> sigma_sales;
  real<lower=0> sigma_scores;
}
model {
  mu_scores ~ normal(pmu_scores, psigma_scores);
  sales ~ normal(mu_sales, sigma_sales);
  scores ~ normal(mu_scores, sigma_scores);
}
generated quantities {
  vector[N] sales_log_lik;
  vector[N] scores_log_lik;
  for (i in 1:N) {
    sales_log_lik[i] = normal_lpdf(sales[i] | mu_sales, sigma_sales);
  }
  for (i in 1:N) {
    scores_log_lik[i] = normal_lpdf(scores[i] | mu_scores, sigma_scores);
  }
}
```

```{stan output.var="sales_hierarchical"}
data {
  int<lower=0> N; // number of data points
  int<lower=0> K; // number of groups
  int<lower=1,upper=K> x[N]; // group indicator
  vector[N] sales; // sales
  vector[N] scores; // critic scores
  real pmu_scores; // prior mean
  real psigma_scores; // prior std
}
parameters {
  vector[K] mu_sales;
  vector[K] mu_scores;
  real<lower=0> sigma_sales;
  real<lower=0> sigma_scores;
}
model {
  mu_scores ~ normal(pmu_scores, psigma_scores);
  sales ~ normal(mu_sales[x], sigma_sales);
  scores ~ normal(mu_scores[x], sigma_scores);
}
generated quantities {
  vector[N] sales_log_lik;
  vector[N] scores_log_lik;
  for (i in 1:N) {
    sales_log_lik[i] = normal_lpdf(sales[i] | mu_sales[x[i]], sigma_sales);
  }
  for (i in 1:N) {
    scores_log_lik[i] = normal_lpdf(scores[i] | mu_scores[x[i]], sigma_scores);
  }
}
```

## Model convergence

We use the potential scale reduction factor $\hat{R}$ for our Stan model’s convergence analysis. We use the improved version proposed by Aki Vehtari
et al. (found at https://arxiv.org/abs/1903.08008).

```{r}
# Separate fit
n_of_groups <- length(unique(data_models$Platform))
n_of_rows <- nrow(data_models)
data_points_sales <- data_models$Global_Sales
data_points_scores <- data_models$Critic_Score

# label each platform by integer for the Stan model
integer_labels <- c("3DS"=1, "DS"=2, "PC"=3, "PS2"=4, "PS3"=5, "PS4"=6, "PSP"=7, "PSV"=8, "Wii"=9, "WiiU"=10, "X360"=11, "XB"=12, "XOne"=13)
group_indicators <- as.numeric(revalue(x=data_models$Platform, integer_labels))

data_group <- list(
  K=n_of_groups,
  sales=data_points_sales,
  scores=data_points_scores,
  N=n_of_rows,
  x=group_indicators,
  pmu_scores=70,
  psigma_scores=0.2
)

separate_fit <- rstan::sampling(sales_separate, data=data_group, iter=1000, chains=4, refresh=0)
```

```{r}
# Hierarchical fit
n_of_groups <- length(unique(data_models$Platform))
n_of_rows <- nrow(data_models)
data_points_sales <- data_models$Global_Sales
data_points_scores <- data_models$Critic_Score

# label each platform by integer for the Stan model
integer_labels <- c("3DS"=1, "DS"=2, "PC"=3, "PS2"=4, "PS3"=5, "PS4"=6, "PSP"=7, "PSV"=8, "Wii"=9, "WiiU"=10, "X360"=11, "XB"=12, "XOne"=13)
group_indicators <- as.numeric(revalue(x=data_models$Platform, integer_labels))

data_group <- list(
  K=n_of_groups,
  sales=data_points_sales,
  scores=data_points_scores,
  N=n_of_rows,
  x=group_indicators,
  pmu_scores=70,
  psigma_scores=0.2
)

hierarchical_fit <- rstan::sampling(sales_hierarchical, data=data_group, iter=1000, chains=4, refresh=0)
```

```{r}
# Pooled fit
n_of_rows <- nrow(data_models)
data_points_sales <- data_models$Global_Sales
data_points_scores <- data_models$Critic_Score

data_group <- list(
  sales=data_points_sales,
  scores=data_points_scores,
  N=n_of_rows,
  pmu_scores=70,
  psigma_scores=0.2
)

pooled_fit <- rstan::sampling(sales_pooled, data=data_group, iter=1000, chains=4, refresh=0)
```

```{r}
print(separate_fit)
```

```{r}
print(hierarchical_fit)
```

```{r}
print(pooled_fit)
```

Each model's  $\hat{R}$ value is sufficiently close to 1. We can therefore assume that each model's chains have reached approximate convergence.

## Model selection

We use the PSIS-LOO -method to asses the predictive performance of the pooled, separate and hierarchical
Gaussian models in the video game sales dataset. Computations are performed on the R PSIS-LOO implementation, package "loo".

### Separate LOO

```{r}
separate_log_lik_sales <- extract_log_lik(separate_fit, parameter_name="sales_log_lik")
separate_rel_n_eff_sales <- relative_eff(exp(separate_log_lik_sales), chain_id=rep(1:4, each=500))
separate_loo_sales <- loo(separate_log_lik_sales, r_eff=separate_rel_n_eff_sales)
separate_log_lik_scores <- extract_log_lik(separate_fit, parameter_name="scores_log_lik")
separate_rel_n_eff_scores <- relative_eff(exp(separate_log_lik_scores), chain_id=rep(1:4, each=500))
separate_loo_scores <- loo(separate_log_lik_scores, r_eff=separate_rel_n_eff_scores)
```

### Pooled LOO

```{r}
pooled_log_lik_sales <- extract_log_lik(pooled_fit, parameter_name="sales_log_lik")
pooled_rel_n_eff_sales <- relative_eff(exp(pooled_log_lik_sales), chain_id=rep(1:4, each=500))
pooled_loo_sales <- loo(pooled_log_lik_sales, r_eff=pooled_rel_n_eff_sales)
pooled_log_lik_scores <- extract_log_lik(pooled_fit, parameter_name="scores_log_lik")
pooled_rel_n_eff_scores <- relative_eff(exp(pooled_log_lik_scores), chain_id=rep(1:4, each=500))
pooled_loo_scores <- loo(pooled_log_lik_scores, r_eff=pooled_rel_n_eff_scores)
```

### Hierarchical LOO

```{r}
hierarchical_log_lik_sales <- extract_log_lik(hierarchical_fit, parameter_name="sales_log_lik")
hierarchical_rel_n_eff_sales <- relative_eff(exp(hierarchical_log_lik_sales), chain_id=rep(1:4, each=500))
hierarchical_loo_sales <- loo(hierarchical_log_lik_sales, r_eff=hierarchical_rel_n_eff_sales)
hierarchical_log_lik_scores <- extract_log_lik(hierarchical_fit, parameter_name="scores_log_lik")
hierarchical_rel_n_eff_scores <- relative_eff(exp(hierarchical_log_lik_scores), chain_id=rep(1:4, each=500))
hierarchical_loo_scores <- loo(hierarchical_log_lik_scores, r_eff=hierarchical_rel_n_eff_scores)
```

```{r}
compare_sales <- loo_compare(x=list(pooled_loo_sales, separate_loo_sales, hierarchical_loo_sales))
compare_scores <- loo_compare(x=list(pooled_loo_scores, separate_loo_scores, hierarchical_loo_scores))
compare_sales
compare_scores
```

We see that the separate model has the largest elpd by a large difference to the pooled and hierarchical models in regards to both of our variables, sales and scores. Thus the PSIS-LOO method strongly suggests that the separate model produces the most reliable results. This is probably because different gaming platforms are too varying in terms of sales and scoring. We proceed with the separate model.

## Separate model posterior visualization

We extract posterior draws of sales and scores for different platforms, compute the ratio of sales to scores and visualize the results.

```{r}
mu_sales <- extract(separate_fit)$mu_sales
mu_scores <- extract(separate_fit)$mu_scores / 100

# c("3DS"=1, "DS"=2, "PC"=3, "PS2"=4, "PS3"=5, "PS4"=6, "PSP"=7, "PSV"=8, "Wii"=9, "WiiU"=10, "X360"=11, "XB"=12, "XOne"=13)
plot_posterior_for_platform <- function(i, label) {
  ratio <- mu_sales[,i] / mu_scores[,i]
  df <- data.frame(ratio=ratio)
  ggplot(df, aes(x=as.numeric(rownames(df)), y=ratio)) + geom_line() + xlab(label)
}

# plot posterior draw ratio for platform
plot_posterior_for_platform(3, "PC")
plot_posterior_for_platform(5, "PS3")
plot_posterior_for_platform(9, "Wii")
plot_posterior_for_platform(2, "DS")

# plot variances for all platforms
variances <- rep(NULL, 13)
platforms <- c("3DS", "DS", "PC", "PS2", "PS3", "PS4", "PSP", "PSV", "Wii", "WiiU", "X360", "XB", "XOne")
for (i in 1:13) {
ratios <- mu_sales[,i] / mu_scores[,i]
  variances[i] <- var(ratios)
}
df <- data.frame(platforms=platforms, variances=variances)
ggplot(df, aes(x=platforms, y=variances)) + geom_col()
```

We find that different platforms have more variance when it comes to the relationship between sales and scores given by critics. It is the platforms developed by Nintendo in particular that have more variance; this could indicate that the consumers for these platforms care less about critic opinions. This could be due to their age, as platforms like Wii and DS have a lot more games made for people of all ages in comparison to a platform like PS3. This is demonstrated in the histograms below.

```{r}
# wii ratings
wii_ratings <- filter(table, Year_of_Release >= 2009, Platform == "Wii")
df.wii <- data.frame(age_rating=wii_ratings$Rating)
ggplot(df.wii, aes(x=age_rating)) + geom_bar() + xlab("Wii")

# DS ratings
# ps3 ratings
ds_ratings <- filter(table, Year_of_Release >= 2009, Platform == "DS")
df.ds <- data.frame(age_rating=ds_ratings$Rating)
ggplot(df.ds, aes(x=age_rating)) + geom_bar() + xlab("DS")

# ps3 ratings
ps3_ratings <- filter(table, Year_of_Release >= 2009, Platform == "PS3")
df.ps3 <- data.frame(age_rating=ps3_ratings$Rating)
ggplot(df.ps3, aes(x=age_rating)) + geom_bar() + xlab("PS3")
```

@bayesian_data_analysis_2013

- prior
- likelihood
- posterior



# References
