---
title: "Bayesian Data Analysis - Project Work"
author: 
- Jaan Tollander de Balsch
- Samuel Piirainen
data: ""
output: 
  pdf_document: 
    latex_engine: xelatex
    toc: yes
    toc_depth: 1
bibliography: bibliography.bib
urlcolor: blue
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/harvard-anglia-ruskin-university.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Subtitle
@bayesian_data_analysis_2013

@video_games_sales

```{r setup, include=FALSE}
library(ggplot2)
library(analogue)
library(tidyr)
library(plyr)
library(dplyr)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

# colClasses drops last column "age rating"
table <- read.csv(
  file="Video_Games_Sales_as_at_22_Dec_2016.csv", header=T, sep=",", fill=T, stringsAsFactors=F, colClasses=c(rep(NA, 15), "NULL"))

# filter out "bad" platforms / years, group by platform
data <- summarise(drop_na(group_by(filter(table, Year_of_Release >= 2010, Platform != "GBA", Platform != "GC"), Platform, Name, Year_of_Release, Global_Sales, Critic_Score)))
```

```{r}
# converts a factor into an integer...
as.numeric.factor <- function(x) {
  as.numeric(levels(x))[x]
}

df.sales <- data.frame(year=data$Year_of_Release, sales=data$Global_Sales)
df.scores <- data.frame(year=data$Year_of_Release, sales=data$Critic_Score)

normalized_sum <- function(arr) {
  return (sum(arr) / length(arr))
}

ggplot(df.sales, aes(x=as.numeric.factor(year), y=sales)) + 
  stat_summary(fun.y = normalized_sum, geom="point", colour = "blue") + geom_smooth(method='lm') + xlab("Year") + ylab("Sales")
```

We normalize the yearly total sales by the length of array of the respective year of sales and see that there's a clear decline in yearly sales.

## Separate model for sales and scores

```{stan output.var="sales_separate"}
data {
  int<lower=0> N; // number of data points
  vector[N] y; // sales
  vector[N] z; // critic scores
}
parameters {
  real mu_y;
  real mu_z;
  real<lower=0> sigma_y;
  real<lower=0> sigma_z;
}
model {
  y ~ normal(mu_y, sigma_y);
  z ~ normal(mu_z, sigma_z);
}
generated quantities {
}
```

```{r}
# sample from the separate model for a platform
sampling_for_platform <- function(data, platformName, iter=1000, chains=4) {
  data <- filter(data, Platform == platformName)
  n_of_rows <- nrow(data)
  data_points_y <- data$Global_Sales
  data_points_z <- data$Critic_Score
  
  data_group <- list(
    N=n_of_rows,
    y=data_points_y,
    z=data_points_z
  )
  
  return (rstan::sampling(sales_separate, data=data_group, iter=iter, chains=chains))
}
```

```{r}
pc_fit <- sampling_for_platform(data, "PC")
ds_fit <- sampling_for_platform(data, "DS")

sales_pc <- extract(pc_fit)$mu_y
scores_pc <- extract(pc_fit)$mu_z
sales_ds <- extract(ds_fit)$mu_y
scores_ds <- extract(ds_fit)$mu_z
```

```{r}
df <- data.frame(sales_pc=sales_pc, scores_pc=scores_pc, sales_ds=sales_ds, scores_ds=scores_ds)
ggplot(df) + geom_point(aes(x=sales_pc, y=scores_pc, colour="PC")) + geom_point(aes(x=sales_ds, y=scores_ds, colour="DS")) +
  scale_colour_manual(values=c("PC" = "red", "DS" = "blue")) + xlab("Sales") + ylab("Scores")
```

## Hierarchical model for sales and scores

```{stan output.var="sales_hierarchical"}
data {
  int<lower=0> N; // number of data points
  int<lower=0> K; // number of groups
  int<lower=1,upper=K> x[N]; // group indicator
  vector[N] y; // sales
  vector[N] z; // critic scores
}
parameters {
  vector[K] mu_y;
  vector[K] mu_z;
  real<lower=0> sigma_y;
  real<lower=0> sigma_z;
}
model {
  y ~ normal(mu_y[x], sigma_y);
  z ~ normal(mu_z[x], sigma_z);
}
generated quantities {
}
```

```{r}
n_of_groups <- length(unique(data$Platform))
n_of_rows <- nrow(data)
data_points_y <- data$Global_Sales
data_points_z <- data$Critic_Score

# label each platform by integer for the Stan model
integer_labels <- c("3DS"=1, "DS"=2, "PC"=3, "PS"=4, "PS2"=5, "PS3"=6, "PS4"=7, "PSP"=8, "PSV"=9, "Wii"=10, "WiiU"=11, "X360"=12, "XB"=13, "XOne"=14)
group_indicators <- as.numeric(revalue(x=data$Platform, integer_labels))

data_group <- list(
  K=n_of_groups,
  y=data_points_y,
  z=data_points_z,
  N=n_of_rows,
  x=group_indicators
)

hierarchical_fit <- rstan::sampling(sales_hierarchical, data=data_group, iter=1000, chains=4)
```

```{r}
mu_y <- extract(hierarchical_fit)$mu_y
mu_z <- extract(hierarchical_fit)$mu_z
ratio_pc <- mu_y[,3] / mu_z[,3]
ratio_ps3 <- mu_y[,6] / mu_z[,6]
ratio_x360 <- mu_y[,12] / mu_z[,12]
ratio_wii <- mu_y[,10] / mu_z[,10]
ratio_ds <- mu_y[,2] / mu_z[,2]
ratio_psv <- mu_y[,9] / mu_z[,9]

df.pc <- data.frame(ratio=ratio_pc)
df.ds <- data.frame(ratio=ratio_ds)
df.x360 <- data.frame(ratio=ratio_x360)
df.ps3 <- data.frame(ratio=ratio_ps3)
df.wii <- data.frame(ratio=ratio_wii)
df.psv <- data.frame(ratio=ratio_psv)

ggplot(df.pc, aes(x=ratio)) + geom_histogram() + xlab("PC Ratio")
ggplot(df.x360, aes(x=ratio)) + geom_histogram() + xlab("X360 Ratio")
ggplot(df.ps3, aes(x=ratio)) + geom_histogram() + xlab("PS3 Ratio")
ggplot(df.wii, aes(x=ratio)) + geom_histogram() + xlab("Wii Ratio")
ggplot(df.ds, aes(x=ratio)) + geom_histogram() + xlab("DS Ratio")
ggplot(df.psv, aes(x=ratio)) + geom_histogram() + xlab("PSV Ratio")

```

Graph seems to suggest that sales correlate with critic scores. Interestingly the correlation varies by platform. The more "hardcore" platforms have less variance in ratio.

```{r}
df <- data.frame(sales_pc=mu_y[,3], scores_pc=mu_z[,3], sales_ds=mu_y[,2], scores_ds=mu_z[,2])
ggplot(df) + geom_point(aes(x=sales_pc, y=scores_pc, colour="PC")) + geom_point(aes(x=sales_ds, y=scores_ds, colour="DS")) +
  scale_colour_manual(values=c("PC" = "red", "DS" = "blue")) + xlab("Sales") + ylab("Scores")
```

The scatterplot also shows how the Nintendo DS sales have far more variance in terms of sales in relation to scores. Sales for Nintendo DS are higher on average, although, again, have more variance than PC games.

# References
